---
title: "Big Rock Pool Challenge - National BioBlitz 2025 – Final Report"
author: "The Rock Pool Project"
output:
  pdf_document:
    latex_engine: xelatex
    includes:
      in_header: report-header.tex
fontsize: 10pt
mainfont: Montserrat
sansfont: Chivo
monofont: Courier New
geometry: a4paper, top=30mm, bottom=20mm, left=20mm, right=20mm
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)

library(ggplot2)
library(dplyr)
library(showtext)
library(ggtext)
library(ggimage)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(knitr)

library(rinat)     # using wrapper for downloading data
library(httr)      # for getting data using inat and WoRMS API's
library(lubridate) # for date conversion
library(jsonlite)  # for getting WoRMS data
library(dplyr)     # for WoRMS data
library(utils)

# Fonts for plots
font_add_google("Montserrat", "mont")
font_add_google("Chivo", "chivo")
showtext_auto()
```

```{r data_load}
load("NatBioBlitz_iNat20250604.RData") #iNat BioBlitz Data as of 4th June 2025
non_native_species <- read.csv("Data/UK marine NNS.csv") # Load the non-native species list
```

\begin{center}
\includegraphics[width=0.8\textwidth]{images/natbioadvert.png}

\includegraphics[width=0.8\textwidth]{images/BRPC Funders and Partner logos.png}
\end{center}

\newpage



# Executive Summary

A brief overview of the National BioBlitz 2025: aims, reach, and key achievements. Include highlights and a general snapshot of participation.

\newpage



# About the National BioBlitz

The [National BioBlitz 2025](https://www.therockpoolproject.co.uk/national-bioblitz) was a UK-wide citizen science initiative coordinated by The Rock Pool Project, aimed at engaging people across the country in discovering and recording the biodiversity of our rocky shores.

Held between 17th and 25th May 2025, the event empowered individuals, families, and wildlife groups to take part in a shared national effort to explore intertidal habitats and contribute meaningful data to marine research and conservation.

Participants of all ages and backgrounds visited their local coastlines to record the species they found in rock pools and along the shore. Whether seasoned naturalists or first-time explorers, everyone played a role in helping to build a clearer picture of the UK's marine life.

## Data collection

All observations were recorded using [iNaturalist](https://www.inaturalist.org/), a free, community-powered biodiversity recording platform. Participants submitted photographs of their finds, along with the date and location of each observation. These were automatically grouped into a dedicated project: the [National BioBlitz 2025 iNaturalist page](https://www.inaturalist.org/projects/brpc-national-bioblitz-2025), which captured all qualifying submissions made during the event period.

Species identifications were suggested by participants and verified through the iNaturalist community and The Rock Pool Project’s team of marine biologists. This process helped ensure data quality while also offering an accessible learning experience for participants.

The resulting dataset represents a unique, community-generated snapshot of intertidal biodiversity across the UK — supporting long-term research and helping to track changes in species distributions, including the spread of marine non-natives.



## Headline Statistics

```{r overall-summary}
# Placeholder for total counts and an icon-based infographic
data_summary <- data.frame(
  category = c("Observers", "Records", "Species"),
  value = c(NA, NA, NA) # Fill with actual numbers
)
```

## Geographical Coverage

```{r geo_coverage}
# Placeholder for map of observations, etc
```

\newpage

# Regional Summaries

Repeat the following structure for each region (e.g. South West, Wales):

## South West Region

- Total observations: _X_
- Unique participants: _Y_
- Most recorded species: _Z_

Individual League table - highlight t shirt winner

```{r regional-plot-sw}
# Placeholder for regional bar chart or map
```

\newpage

# Individual Feedback (optional)

Tailor info for a specific individual

## Example User Summary

- Total records: _X_
- Species recorded: _Y_
- Non-native species: _Z_

top species

```{r individual-feedback-example}
# Placeholder for individual bar/tile chart
```

\newpage

# Non-native Species Results

## Summary from the National BioBlitz

### Overview of the non-native species found

#### Taxonomic breakdown of native and non-native species

```{r nn-summary}
# Placeholder for top non-native species chart
```


```{r bioblitzdata,fig.align = "center",fig.width = 8,fig.height = 5}
# Define project ID and API parameters
project_slug <- "brpc-national-bioblitz-2025-practice"

# Download data using the rinat package
inat_data <- get_inat_obs_project(project_slug)

# Convert observed_on to date-time for comparison
inat_data$updated_at <- ymd_hms(inat_data$updated_at)
inat_data$time_observed_at <- ymd_hms(inat_data$time_observed_at)

# Create empty columns to bind with the inat_data dataframe for the new taxonomic data
taxon.kingdom = rep(NA, times = length(inat_data[,1]))
taxon.phylum = rep(NA, times = length(inat_data[,1]))
taxon.class = rep(NA, times = length(inat_data[,1]))
taxon.order = rep(NA, times = length(inat_data[,1]))
taxon.family = rep(NA, times = length(inat_data[,1]))
marine = rep(NA, times = length(inat_data[,1]))
brackish = rep(NA, times = length(inat_data[,1]))
freshwater = rep(NA, times = length(inat_data[,1]))
terrestrial = rep(NA, times = length(inat_data[,1]))

# Combine the inat_data dataframe with the new columns
inat_data = cbind(
  inat_data,taxon.kingdom,taxon.phylum,taxon.class,taxon.order,taxon.family,
  marine,brackish,freshwater,terrestrial)

# Fill in new inat_data columns with taxonomic information from WoRMS
l = length(inat_data[,1])
for (a in 1:l) {
  if (
    is.na(inat_data$taxon.rank[a])==FALSE && inat_data$taxon.rank[a]=="species"
    ){
    
    # Split the scientific name into two character objects: genus and species
    binomClassNm = inat_data$taxon.name[a]
    binomClassNmSplit = strsplit(binomClassNm,"[ ]")
    
    genus = binomClassNmSplit[[1]][1]
    species = binomClassNmSplit[[1]][2]
    
    # Paste the genus and species names into the WoRMS API key and download relevant data
    api = paste("https://www.marinespecies.org/rest/AphiaRecordsByName/",
                genus,
                "%20",
                species,
                "?like=true&marine_only=false&extant_only=true&offset=1",
                sep = "")
    taxonInfo = GET(api)
    taxonInfoContent = httr::content(taxonInfo, as = 'text')
    
    # Populate the new columns with taxonomic data downloaded from WoRMS
    if(object.size(taxonInfoContent)>112) {
      taxonInfoContentJSON = jsonlite::fromJSON(taxonInfoContent)
      
      inat_data$taxon.kingdom[a] = taxonInfoContentJSON$kingdom[1]
      inat_data$taxon.phylum[a] = taxonInfoContentJSON$phylum[1]
      inat_data$taxon.class[a] = taxonInfoContentJSON$class[1]
      inat_data$taxon.order[a] = taxonInfoContentJSON$order[1]
      inat_data$taxon.family[a] = taxonInfoContentJSON$family[1]
      
      if(is.na(taxonInfoContentJSON$isMarine[1])==FALSE) {
        inat_data$marine[a] = taxonInfoContentJSON$isMarine[1]
      }
      if(is.na(taxonInfoContentJSON$isBrackish[1])==FALSE) {
        inat_data$brackish[a] = taxonInfoContentJSON$isBrackish[1]
      }
      if(is.na(taxonInfoContentJSON$isFreshwater[1])==FALSE) {
        inat_data$freshwater[a] = taxonInfoContentJSON$isFreshwater[1]
      }
      if(is.na(taxonInfoContentJSON$isTerrestrial[1])==FALSE) {
        inat_data$terrestrial[a] = taxonInfoContentJSON$isTerrestrial[1]
      }
    } else {
      inat_data$taxon.kingdom[a] = "taxon info not retrieved"
    }
  }
}
# Match observations against non-native species list
natbioblitz_nns <- subset(inat_data, taxon.id  %in% non_native_species$inat_id)

########## plot #################
a=2
#-------------------------------------------------------------------------------
# Sort out data for bar plots
#-------------------------------------------------------------------------------

# Create the bar plots
# Filter out the non-terrestrial species data and species not identified to species level
inat_data_filtered = inat_data
inat_data_filtered = filter(inat_data_filtered,
                            marine == 1 | brackish == 1 | freshwater ==1)

# Define column reference number for the name of each taxa group for each of the three ranks
colRefINat_data_filtered=length(inat_data_filtered)-(5+a)

# Sort each of the names alphabetically to allow accurate matching when native and non-native species tables are combined
taxonNames = sort(unique(
  inat_data_filtered[,colRefINat_data_filtered]))

# Create empty matrix for bar plot data
df = matrix(0,nrow=1,ncol=length(taxonNames))
colnames(df)=taxonNames
rownames(df)=c("nonNative")

# Define column reference number for the name of each taxa group
colRefNatbioblitz_nns = length(natbioblitz_nns)-(5+a)

# Create frequency tables for native and non-native species
nativeTable = table(inat_data_filtered[,colRefINat_data_filtered])
nonNativeTable = table(natbioblitz_nns[,colRefNatbioblitz_nns])

# Populate the empty matrix for bar plot data with frequency values of each non-native species from the non-native species frequency table
for (b in 1:ncol(df)){
  for (c in 1:length(dimnames(nonNativeTable)[[1]])){
    if (colnames(df)[b] == dimnames(nonNativeTable)[[1]][c]){
      df[b] = nonNativeTable[c]
    }
  }
}

# Re-define the matrix object as having the rows from the native species frequency table bound to the rows of the matrix (now populated with non-native species frequencies of occurrence)
df = rbind(nativeTable,df)

# Subtract the non-native species frequencies from the native species frequency row
for (d in 1:ncol(df)) {
  df[1,d] = df[1,d]-df[2,d]
}

#-------------------------------------------------------------------------------
# Plot the bar plot
#-------------------------------------------------------------------------------
op = par(mfrow=c(1,3),family = "mont", font.lab = 2,
         mar=c(5,8,2,2)+0.1,xpd=FALSE)
barplot(df, 
        col=c("#00a6fb","#F79824"), 
        horiz = TRUE, cex.names = 0.9,las = 1,border = "#FFFFFF", 
        space=0.4,
        font.axis=1,
        cex.axis = 0.9,
        cex.lab = 0.9, 
        xlab="Number of records",
        col.lab =c("#191d2d"),family="mont"
        )
# Legend
legend("bottomright", inset = c(0.1),
       fill = c("#00a6fb","#F79824"),
       legend=c("Native species",
                "Non-native species"), horiz = FALSE,cex = 0.9,border = FALSE)
# Outer plot title
mtext("Number of non-terrestrial records identified to species level",
      side = 3, line = -1.5, outer = TRUE,col = c("#191d2d"),font = 2,cex = 1.1,
      family="chivo")
```

## Comparison with NBN Atlas Baseline

 - Which species were found in unusual places?
 - Which species weren't found in places they would be expected to be found?

```{r nn-comparison,fig.align = "center",fig.width = 8,fig.height = 5}

################################################################################
#                 Maps of records from the Bioblitz and NBN Atlas
################################################################################

# Download the base map of UK and surrounding countries
uk_map <- ne_countries(
  scale = "medium",
  returnclass = "sf"
) %>%
  filter(admin %in% c("United Kingdom", "Ireland",
                      "Jersey","France","Netherlands","Germany",
                      "Denmark","Belgium","Norway","Finland"))

# String of species names to loop through
nonNatives = unique(natbioblitz_nns$taxon.name)
nonNativesCN = unique(natbioblitz_nns$taxon.common_name.name)

# Set graphical parameters for the maps
op = par(mfrow=c(2,4),family = "mont", font.lab = 2,
         mar=c(2,2.5,1,0.1)+0.1,
         oma=c(0.01,0.01,2,0.01),xpd=FALSE)

###### Get NBN Atlas data and map the records from bioblitz and NBN Atlas ######
for (a in 1:length(nonNatives)) {
  
  # Split species name into genus and species to parse the API for NBN Atlas
  binomClassNm = nonNatives[a]
  binomClassNmSplit = strsplit(binomClassNm,"[ ]")
  genus = binomClassNmSplit[[1]][1]
  species = binomClassNmSplit[[1]][2]
  
  # NBN Atlas record filter:
  #     - No unconfirmed, unconfirmed (not reviewed) or unconfirmed (plausible)
  #     - No absences
  api=paste("https://records-ws.nbnatlas.org/occurrences/search?",
            "q=*:*&fq=genus:",
            genus,
            "&fq=-(identification_verification_status%3A%22Unconfirmed%22%20OR",
            "%20identification_verification_status%3A%22Unconfirmed%20-%20",
            "not%20reviewed%22%20OR%20identification_verification_status%3A%22",
            "Unconfirmed%20-%20plausible%22)&fq=-occurrence_status%3A%22absent",
            "%22&fq=taxon_name%3A%22",
            genus,
            "%20",
            species,
            "%22&pageSize=10000",sep = "")
  taxonInfo = GET(api)
  taxonInfoContent = httr::content(taxonInfo, as = 'text')
  taxonInfoContentJSON = jsonlite::fromJSON(taxonInfoContent)
  df=taxonInfoContentJSON$occurrences
  
  #################### Filter the records from bioblitz ########################
  bioblitzFiltered = subset(natbioblitz_nns,
                            taxon.name == binomClassNm)
  
  ############### Map the records from bioblitz and NBN Atlas ##################
  
  # Projection/CRS of base map, NBN Atlas coordinates and Bioblitz coordinates:
  #     - WGS 84
  plot(st_geometry(uk_map),border="#FFFFFF",axes=TRUE,
       xlim=c(-15,5),ylim=c(48.5,61.5),
       family="mont",col="#E4E6F6",cex.axis=0.8)
  points(as.numeric(df$decimalLongitude),
         as.numeric(df$decimalLatitude),
         pch = 19,
         cex = 0.7,
         col="#00A6FB")
  points(as.numeric(bioblitzFiltered$longitude),
         as.numeric(bioblitzFiltered$latitude),
         pch = 19,
         cex = 0.9,
         col="#F79824")
  title(main = nonNativesCN[a],family="mont",cex.main = 0.9,line = -1)
  legend("topleft", ,inset = c(0.02,0.1),
         col = c("#00a6fb","#F79824"),
         legend=c("NBN Atlas",
                  "Bioblitz"),
         horiz = FALSE,cex = 0.7,pch=19,box.lwd=0.01,box.col = "white")
  mtext("Non-native Invasive Species Records",
        side = 3, line = -0.14, outer = TRUE,col = c("#191d2d"),
        font = 2,cex = 1.05,family="chivo")
  rm(taxonInfo,taxonInfoContent,taxonInfoContentJSON,df)
}
```

\newpage

# Species Highlights

- Rarest record
- Best photo submission
- Noteworthy range extension

```{r highlights}
# Placeholder for image collage or featured species panel
```


\newpage

# Project Partners and Support

Thank you to all our partners, volunteers, and funders.

ADD PARTNER LOGOS HERE

\newpage

# Next Steps: How to Stay Involved

## 1. Join or Start a Rock Pool Hub
- Get involved locally or start your own hub with our support.

## 2. Keep Recording on iNaturalist
- Every observation adds to the UK marine picture.

## 3. Seasonal Challenges
- Look out for themed events and mini-BioBlitzes.

## 4. Follow & Tag Us
- Stay connected via social media: [@therockpoolproject](https://www.instagram.com/therockpoolproject)

## 5. Volunteer or Collaborate
- Run events, support schools, or partner with us.

## 6. Sign Up for Updates
- Visit [www.therockpoolproject.co.uk](https://www.therockpoolproject.co.uk)

\newpage

# Appendices

## Species List

```{r species-table}
# Placeholder for table of all recorded species
```

## Methodology Notes

Details on data collection, validation, and tools used.

## Data Use Statement

Explanation of how participant and observation data are used, stored, and shared.
